---
title: "Obama Tweets"
description: |
  Analysis of Barack Obama’s archived tweets, focusing on hashtags and pronoun usage.
author: Colby Cooley
date: October 6, 2025
format: html
execute: 
  warning: false
  message: false
---

## Introduction

This analysis uses archived tweets from Barrack Obama's Twitter page to explore two aspects of the former president’s communication on Twitter. First, the hashtags he used most frequently, and second, his use of collective versus individual pronouns. Tweets are taken from the publicly released Obama Administration social media archive, which includes all tweets posted by @POTUS (Obama) during his presidency. The dataset provides a useful lens into how he framed political messaging.

```{r}
library(tidyverse)

obama_tweets <- 
  read_csv(unz("data/POTUS111716.zip",
               "tweets.csv")) |>
  select(text, timestamp) |>
  
  # extract hashtags
  mutate(hashtags = str_extract_all(text,
                                    "(?<=#)\\w+")) |>
  
  # clean text and remove URLs/usernames
  mutate(text_clean = str_replace_all(text,
                                      "https?://\\S+|@\\w+", 
                                      ""), 
         text_clean = str_to_lower(text_clean)) |> 
  
  # count collective and individual pronouns
  mutate(collective_count = str_count(text_clean, 
                                      "\\b(we|us|our|ours|we're|we've)\\b"),
         individual_count = str_count(text_clean, 
                                      "\\b(i|me|my|mine|i've|i'm)\\b"))
```

## Most Common Hashtags

First, we must extract and count all of his hashtags.

```{r}
obama_hashtags <- 
  obama_tweets |> 
  select(hashtags) |> 
  unnest(hashtags) |> 
  group_by(hashtags) |> 
  summarize(count = n()) |> 
  arrange(desc(count)) |> 
  head(5)
```

Then, we can visualize the data.

```{r}
#| fig-alt: "Bar chart showing Barack Obama's five most frequently used hashtags on Twitter."
#| fig-cap: "Obama’s five most frequently used hashtags."

ggplot(obama_hashtags, 
       aes(x = hashtags, 
           y = count, 
           fill = hashtags)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Frequency of Obama's Top Five Most Used Twitter Hashtags", 
       x = "Hashtag", 
       y = "Frequency")
```

The plot reveals which issues he emphasized publicly and illustrates some of the ways that the former president used Twitter to engage with the general public.

## Collective vs. Individual Pronoun Usage

First, we must get the totals for each type of pronoun.

```{r}
total_pronouns <- obama_tweets |>
  summarise(
    collective_total = sum(collective_count),
    individual_total = sum(individual_count)
  ) |>
  pivot_longer(cols = everything(), names_to = "pronoun_type", values_to = "total")
```

Then, we can visualize the data.

```{r}
#| fig-alt: "Bar chart comparing Obama's usage of collective pronouns versus individual pronouns in his tweets."
#| fig-cap: "Comparison of collective vs. individual pronoun usage in Obama’s tweets."

ggplot(total_pronouns, aes(x = pronoun_type, y = total, fill = pronoun_type)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Total Usage of Collective vs. Individual Pronouns",
    x = "Pronoun Type",
    y = "Total Count"
  ) +
  theme_minimal()
```

The plot shows how Obama uses collective pronouns much more often than individual ones, and gives us insight into how Obama purposefully focused his tweets on unity and shared identity rather than on himself.

## References

National Archives and Records Administration. (n.d.). *Obama White House Social Media Archive: Tweets.*

Retrieved from the Barack Obama Presidential Library:

<https://www.obamalibrary.gov/research/obama-white-house-social-media-archive>
