---
title: "Obama Tweets"
description: |
  Analysis of Barack Obama's Tweets
author: Colby Cooley
date: October 6, 2025
format: html
execute: 
  warning: false
  message: false
---

Data from: National Archives and Records Administration. (n.d.). Obama White House social media archive: Tweets. Retrieved from Obama Federal Archives: Barack Obama Presidential Library. 
obamalibrary.gov

```{r}
library(tidyverse)
```

```{r}
obama_tweets <- read_csv(unz("data/POTUS111716.zip", "tweets.csv"))
obama_tweets <- obama_tweets |> select(text, timestamp)
```

```{r}
obama_tweets <- obama_tweets |> mutate(hashtags = str_extract_all(text, "(?<=#)\\w+"))
```

```{r}
obama_hashtags <- obama_tweets |> select(hashtags) |> unnest(hashtags) |> group_by(hashtags) |> summarize(count = n()) |> arrange(desc(count)) |> head(5)
```

The following plot shows Obama's most used hashtags, revealing which issues he emphasized publicly. It also illustrates some of the ways that the former president used Twitter to engage with the general public. 

```{r}
ggplot(obama_hashtags, aes(x = hashtags, y = count, fill = hashtags)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Frequency of Obama's Top Five Most Used Twitter Hashtags", x = "Hashtag", y = "Frequency")
```

```{r}
obama_tweets <- obama_tweets |> 
  mutate(text_clean = str_replace_all(text, "https?://\\S+|@\\w+", ""), text_clean = str_to_lower(text_clean))
```

```{r}
obama_tweets <- obama_tweets |> mutate(collective_count = str_count(text_clean, "\\b(we|us|our|ours|we're|we've)\\b"), individual_count = str_count(text_clean, "\\b(i|me|my|mine|i've|i'm)\\b"))
```

```{r}
obama_pronouns <- obama_tweets |> pivot_longer(c(collective_count, individual_count), names_to = "pronoun_type", values_to = "count")
```

The following plot shows the distribution of collective pronouns(i.e. we/us/our) in Obama's tweets as well as the distribution of individual pronouns(i.e. I/me/my) in Obama's tweets. It gives us insight into how Obama purposefully focused his tweets on unity and shared identity rather than on himself.

```{r}
ggplot(obama_pronouns, aes(x = count)) +
  facet_grid("pronoun_type", labeller = labeller(pronoun_type = c(collective_count = "Collective",
  individual_count = "Individual"))) +
  geom_boxplot() +
  scale_x_log10() +
  labs(title = "Distribution of Collective vs. Individual Pronouns in Obama's Tweets (Log Scale)", x = "Pronoun Count per Tweet (log scale)")
```

